---
title: "STOR 664 — Part 1 Combined EDA (Pokémon)"
author: "Team: Yinyu Yao, Kareena Legare, Samuel Moore, Irene Zhang"
date: "2025-11-15"
output:
  pdf_document: default
  html_document:
    toc: true
    toc_float: true
urlcolor: blue
---
## Introduction
We used the Pokemon dataset from TidyTuesday. [Dataset Link](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-04-01/readme.md) 

Pokemon is a Japanese video game where you catch Pokemon in the wild and use them to fight your battles. Pokemon are wild monsters and each one has a type, like fire, water, grass, etc. They also have basic stats like HP, attack, defense etc. The dataset we chose 949 rows and 22 columns with information about the Pokemon's name, height weight, stats, type etc. 

The question we want to answer is: What characteristics of Pokemon best predict their HP stat? Is there a relationship between HP and the Pokemon's other characteristics? 

There are many existing projects that explore a Pokemon dataset. These can be found on Kaggle and Medium. Many of them do data visualization or EDA. In particular, this [Project](https://medium.com/@hanahshih46/pokemon-data-visualization-and-analysis-with-r-60970c8e37f4) attempts to answer a similar question as ours and performed a simple linear regression with HP as their Y-variable. 

## EDA
We first did extensive EDA on our dataset. 

```{r setup, include=FALSE}
# Load Packages and Data

library(tidytuesdayR)
library(tidyverse)
library(janitor)
library(forcats)
library(broom)
library(readr)
library(patchwork)
library(corrplot)

# Keep dirs 
dir.create("data/processed", recursive = TRUE, showWarnings = FALSE)
dir.create("tables",         recursive = TRUE, showWarnings = FALSE)
dir.create("figures",        recursive = TRUE, showWarnings = FALSE)

tues <- tidytuesdayR::tt_load(2025, week = 13)
pokemon <- tues$pokemon_df |>

select(-id, -species_id, -color_1, -color_2, -color_f, -url_icon, -url_image) |>
clean_names() |>
mutate(
dual_type       = !is.na(type_2),
log_weight      = if_else(weight > 0, log(weight), NA_real_),
log_height      = if_else(height > 0, log(height), NA_real_),
base_stat_total = hp + attack + defense + special_attack + special_defense + speed
)

write_csv(pokemon, "data/processed/pokemon_processed.csv")


```

```{r}
# 1)  Dataset Summary and Statistics

# NA counts across ALL columns (1-row audit table)
na_counts <- pokemon |>
  summarise(across(everything(), ~sum(is.na(.)))) |>
  tidyr::pivot_longer(
    everything(),
    names_to  = "variable",
    values_to = "na_count"
  ) |>
  dplyr::filter(na_count > 0) |>
  arrange(desc(na_count))
knitr::kable(na_counts, digits = 0,
             caption = "Columns with at least one missing value.")
 
# Type1 / Generation counts
gen_tab  <- as.data.frame(table(pokemon$generation_id, useNA = "ifany"))
names(gen_tab) <- c("generation_id","count")
type_tab <- as.data.frame(table(pokemon$type_1,       useNA = "ifany"))
names(type_tab) <- c("type_1","count")

knitr::kable(gen_tab,  caption = "Counts of Pokémon by generation_id")
knitr::kable(type_tab, caption = "Counts of Pokémon by type_1")

# egg_group_1 counts
egg1_counts <- pokemon |> count(egg_group_1) |> arrange(desc(n))
knitr::kable(egg1_counts, caption = "Counts of egg_group_1")

```

```{r}
# 2) Conditional distribution P(type_2 | type_1) + X² ----

pokemon_types <- pokemon |> filter(!is.na(type_2))

type_pair_counts <- pokemon_types |>
count(type_1, type_2, name = "n") |>
group_by(type_1) |>
mutate(row_total = sum(n), prop = n / row_total) |>
arrange(type_1, desc(prop)) |>
ungroup()

knitr::kable(head(type_pair_counts, 15), digits = 3,
caption = "P(type_2 | type_1): top 15 rows.")

type2_pref <- type_pair_counts |>
group_by(type_1) |>
slice_max(prop, n = 1, with_ties = TRUE) |>
arrange(type_1, desc(prop)) |>
ungroup()

knitr::kable(type2_pref, digits = 3,
caption = "Most frequent secondary type(s) for each primary type.")

tab_t1_t2 <- table(type_1 = pokemon_types$type_1, type_2 = pokemon_types$type_2)
chi_res <- chisq.test(tab_t1_t2)  # prints Pearson X²
chi_tbl <- tibble(statistic = as.numeric(chi_res$statistic),
df        = as.integer(chi_res$parameter),
p_value   = as.numeric(chi_res$p.value))
knitr::kable(chi_tbl, caption = "Chi-squared test of independence between type_1 and type_2.")
chi_res

```

```{r}
# 3) By generation_id: distribution tables and boxplots

pokemon <- pokemon |>
mutate(generation_id_f = forcats::fct_na_value_to_level(as.factor(generation_id), level = "Unknown"))
stat_vars <- c("hp","attack","defense","special_attack","special_defense","speed")

gen_stats_all <- pokemon |>
  group_by(generation_id_f) |>
  summarise(
    n = n(),
    across(all_of(stat_vars),
           list(mean = ~mean(.x, na.rm = TRUE),
                sd   = ~sd(.x,   na.rm = TRUE)),
           .names = "{.col}_{.fn}")
  )

gen_stats_long <- gen_stats_all |>
  tidyr::pivot_longer(
    cols = -c(generation_id_f, n),
    names_to = c("stat", "metric"),
    names_sep = "_(?=[^_]+$)"
  ) |>
  tidyr::pivot_wider(
    names_from  = metric,
    values_from = value
  ) |>
  # ensure stat is ordered like stat_vars: hp, attack, defense, ...
  mutate(stat = factor(stat, levels = stat_vars)) |>
  arrange(stat, generation_id_f)

knitr::kable(
  gen_stats_long,
  digits = 1,
  caption = "Means and SDs of stats by generation (including Unknown)."
)

gen_plots <- lapply(stat_vars, function(s) {
  ggplot(pokemon, aes(x = generation_id_f, y = .data[[s]])) +
    geom_boxplot(outlier.alpha = 0.4) +
    labs(
      title = paste("Dist. of", s, "by generation"),
      x = "Generation",
      y = s
    ) +
    theme_bw()
})
names(gen_plots) <- stat_vars

gen_panel_1 <- gen_plots[["hp"]]             + gen_plots[["attack"]]
gen_panel_2 <- gen_plots[["defense"]]        + gen_plots[["special_attack"]]
gen_panel_3 <- gen_plots[["special_defense"]] + gen_plots[["speed"]]

gen_panel_1
gen_panel_2
gen_panel_3

```


```{r}
# 4) By type1: distribution tables and boxplots

pokemon <- pokemon |>
  mutate(type_1_f = forcats::fct_infreq(as.factor(type_1)))

type1_stats <- pokemon |>
  group_by(type_1_f) |>
  summarise(
    n = n(),
    across(all_of(stat_vars),
           list(mean = ~mean(.x, na.rm = TRUE),
                sd   = ~sd(.x,   na.rm = TRUE)),
           .names = "{.col}_{.fn}")
  )

# Table 1: HP and Speed
type1_hp_speed <- type1_stats |>
  select(type_1_f, n, hp_mean, hp_sd, speed_mean, speed_sd)

knitr::kable(
  type1_hp_speed,
  digits = 1,
  caption = "Means and SDs of HP and Speed by primary type (type_1)."
)

# Table 2: Attack and Special Attack
type1_atk_spa <- type1_stats |>
  select(type_1_f, n,
         attack_mean, attack_sd,
         special_attack_mean, special_attack_sd)

knitr::kable(
  type1_atk_spa,
  digits = 1,
  caption = "Means and SDs of Attack and Special Attack by primary type (type_1)."
)

# Table 3: Defense and Special Defense
type1_def_spdef <- type1_stats |>
  select(type_1_f, n,
         defense_mean, defense_sd,
         special_defense_mean, special_defense_sd)

knitr::kable(
  type1_def_spdef,
  digits = 1,
  caption = "Means and SDs of Defense and Special Defense by primary type (type_1)."
)


type_plots <- lapply(stat_vars, function(s) {
  ggplot(pokemon, aes(x = type_1_f, y = .data[[s]])) +
    geom_boxplot(outlier.alpha = 0.4) +
    coord_flip() +
    labs(
      title = paste("Dist. of", s, "by type_1"),
      x = "Primary type (type_1)",
      y = s
    ) +
    theme_bw()
})
names(type_plots) <- stat_vars

type_panel_1 <- type_plots[["hp"]]             + type_plots[["attack"]]
type_panel_2 <- type_plots[["defense"]]        + type_plots[["special_attack"]]
type_panel_3 <- type_plots[["special_defense"]] + type_plots[["speed"]]

type_panel_1
type_panel_2
type_panel_3

```


```{r}
# 5) Size → HP (raw vs log)

p_hp_h  <- ggplot(pokemon, aes(height, hp))      + geom_point() + theme_bw() + labs(title="HP vs height")
p_hp_lh <- ggplot(pokemon, aes(log(height), hp)) + geom_point() + theme_bw() + labs(title="HP vs log(height)")
p_hp_w  <- ggplot(pokemon, aes(weight, hp))      + geom_point() + theme_bw() + labs(title="HP vs weight")
p_hp_lw <- ggplot(pokemon, aes(log(weight), hp)) + geom_point() + theme_bw() + labs(title="HP vs log(weight)")

hp_panel <- (p_hp_h + p_hp_lh) /
            (p_hp_w + p_hp_lw)
hp_panel

```

```{r}
# 6) Variable Relations

p_atk_spa <- ggplot(pokemon, aes(attack, special_attack))  + geom_point() + theme_bw() + labs(title="Special Attack vs Attack")
p_def_spd <- ggplot(pokemon, aes(defense, special_defense)) + geom_point() + theme_bw() + labs(title="Special Defense vs Defense")
atk_def_panel <- p_atk_spa + p_def_spd
atk_def_panel

eq_counts <- tibble(
attack_eq_spatk  = sum(pokemon$attack  == pokemon$special_attack,  na.rm = TRUE),
defense_eq_spdef = sum(pokemon$defense == pokemon$special_defense, na.rm = TRUE),
both_equal       = sum((pokemon$attack  == pokemon$special_attack) &
(pokemon$defense == pokemon$special_defense), na.rm = TRUE)
)
knitr::kable(eq_counts, digits = 3)

```

```{r}
# 7) Dual-type densities

p_den_hp  <- ggplot(pokemon, aes(hp,             color = dual_type)) + geom_density() + theme_minimal() + labs(title="HP by dual_type", y=NULL)
p_den_def <- ggplot(pokemon, aes(defense,        color = dual_type)) + geom_density() + theme_minimal() + labs(title="Defense by dual_type", y=NULL)
p_den_exp <- ggplot(pokemon, aes(base_experience,color = dual_type)) + geom_density() + theme_minimal() + labs(title="Base experience by dual_type", y=NULL)

density_panel <- p_den_hp / p_den_def / p_den_exp
density_panel

dual_means_atk <- pokemon |> group_by(dual_type) |> summarise(`mean(attack)` = mean(attack, na.rm=TRUE), .groups="drop")
knitr::kable(dual_means_atk, digits = 3)

```

```{r}
# 8) Correlation heatmap (numeric cols excluding generation_id)

num_keep <- pokemon |> select(where(is.numeric)) |> select(!generation_id)
M <- cor(num_keep, use = "pairwise.complete.obs")
corrplot::corrplot(M, method = "color", type = "upper", addCoef.col = "black", tl.col = "black",
number.cex = 0.5, tl.srt = 45)

png("figures/k_corrplot_matrix.png", width = 1200, height = 1200, res = 140)
corrplot::corrplot(M, method = "color", type = "upper", addCoef.col = "black", tl.col = "black",
number.cex = 0.5, tl.srt = 45)
dev.off()

```

```{r}
# 9) base_stat_total by generation

gen_total <- pokemon |>
group_by(generation_id_f) |>
summarise(n = n(),
mean_total = mean(base_stat_total, na.rm=TRUE),
sd_total   = sd(base_stat_total,   na.rm=TRUE)) |>
arrange(generation_id_f)

knitr::kable(gen_total, digits = 1)

p_gen_total <- ggplot(pokemon, aes(generation_id_f, base_stat_total)) +
geom_boxplot(outlier.alpha=.4) + theme_bw() +
labs(title="Base stat total by generation (incl. Unknown)", x="Generation", y="Base stat total")
p_gen_total

```


```{r}
# 10) Speed by dual_type (boxplot + quick t-test)

p_speed_dual <- ggplot(pokemon, aes(x = dual_type, y = speed)) +
geom_boxplot(outlier.alpha=.4) + theme_bw() +
labs(title="Speed by dual_type", x="Dual type?", y="Speed")
p_speed_dual

tt_speed <- broom::tidy(t.test(speed ~ dual_type, data = pokemon))
knitr::kable(tt_speed, digits = 3)

```

```{r}
# 11) Base experience VS base stat total (scatter + OLS line)

p_exp_total <- ggplot(pokemon, aes(base_experience, base_stat_total)) +
geom_point(alpha=.6) + geom_smooth(method="lm", se=FALSE) + theme_bw() +
labs(title="Base experience vs base stat total", x="Base experience", y="Base stat total")
ggsave("figures/baseexp_vs_basetotal.png", p_exp_total, width = 7, height = 5, dpi = 150)
p_exp_total

```

```{r}
# 12) Speed vs log(weight), faceted by top 6 primary types

top6 <- pokemon |> count(type_1) |> arrange(desc(n)) |> slice_head(n=6) |> pull(type_1)
p_speed_lw_facets <- pokemon |>
filter(type_1 %in% top6, !is.na(log_weight)) |>
ggplot(aes(log_weight, speed)) +
geom_point(alpha=.55) + geom_smooth(method="lm", se=FALSE) +
facet_wrap(~type_1, ncol=3) + theme_bw() +
labs(title="Speed vs log(weight): top 6 primary types", x="log(weight)", y="Speed")
ggsave("figures/speed_vs_logweight_top6_types.png", p_speed_lw_facets, width = 9, height = 6, dpi = 150)
p_speed_lw_facets

```


### Null Hypotheses
After doing EDA, we determined some potential null hypotheses that are worth further exploring. 

$$
H_0^{(1)}: \beta_{\text{speed}} = \beta_{\text{attack}} = 0, \quad
H_0^{(2)}: \beta_{\text{dual\_type\_x\_defense}} = 0, \quad
H_0^{(3)}: \beta_{\text{base\_experience}} = 0
$$

### Plots Motivating our Hypothesis
```{r}
#1. 
p1 <- ggplot(pokemon, aes(speed, hp))     + geom_point() + geom_smooth(method='lm', se=FALSE) + theme_minimal() + labs(title="HP vs. Speed")
p2 <- ggplot(pokemon, aes(attack, hp))    + geom_point() + geom_smooth(method='lm', se=FALSE) + theme_minimal() + labs(title="HP vs. Attack")
p3 <- ggplot(pokemon, aes(speed, attack)) + geom_point() + geom_smooth(method='lm', se=FALSE) + theme_minimal() + labs(title="Attack vs. Speed")
g_triptych <- p1 + p2 + p3 + patchwork::plot_layout(ncol=2)
ggsave("figures/k_hypothesis_triptych.png", plot = g_triptych, width = 9, height = 6, dpi = 150)
g_triptych

corrs_simple <- pokemon |>
summarise(`cor(speed, hp)` = cor(speed, hp),
`cor(speed, attack)` = cor(speed, attack),
`cor(attack, hp)` = cor(attack, hp))
knitr::kable(corrs_simple, digits = 3)
```

```{r}
#2
ggplot(pokemon) +
geom_density(aes(x = defense, color = dual_type)) +
theme_minimal() +
labs(title = "Distribution of Defense Score by Dual Type",
color = "Dual Type",
x = "Defense",
y = "")

```

```{r}
#3
ggplot(pokemon) +
geom_point(aes(x = base_experience, y = hp)) +
labs(title = "HP vs. Base Experience",
x = "Base Experience",
y = "HP") +
theme_minimal()

#special attack/defense motivation
special_df <- pokemon |>
  mutate(
    same_attack  = special_attack  == attack,
    same_defense = special_defense == defense
  )

p_same_atk <- ggplot(special_df) +
  geom_point(aes(x = attack, y = special_attack, color = same_attack)) +
  labs(
    title = "Identical Values of Special Attack and Attack",
    x = "Attack",
    y = "Special Attack",
    color = "Identical Values"
  ) +
  theme_minimal()

p_same_def <- ggplot(special_df) +
  geom_point(aes(x = defense, y = special_defense, color = same_defense)) +
  labs(
    title = "Identical Values of Special Defense and Defense",
    x = "Defense",
    y = "Special Defense",
    color = "Identical Values"
  ) +
  theme_minimal()

same_panel <- p_same_atk / p_same_def
same_panel


```

### Limitations of dataset
This dataset doesn't include some characteristics of Pokemon that may or may not be helpful in answering our questions. For example, it doesn't include the catch rate of Pokemon or if the Pokemon is legendary/mythical or not. 

## Proposed analysis plan
The team will carry out an iterative, back-and-forth model-building process within the linear regression framework for predicting HP, trying different combinations of predictors and interactions. At each stage, we will use methods discussed in class—primarily *t-tests* for individual coefficients and *F-tests* for groups of terms—to assess which variables and interactions meaningfully contribute to the model. Once we arrive at a well-supported specification, we will use the fitted model to test the three hypotheses outlined earlier, reject or fail to reject them based on the results, and summarize our conclusions in the final project write-up.

## Acknowledgements 
ChatGPT was used to help with coding. Yin and Kareena did the EDA and made the figures. Irene did the written portion of the report. Sam made the presentation.
